---
title: "R-Introduction using RStudio"
author: "S. Ravichandran, Ph.D"
date: "Jan 21, 2020; updated 01/2022"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### History of R 

* R is a language 
* No compilation
* Dynamic programming language

### RStudio basics 

* RStudio is a RGUI
* R is the main program 

Helpful to start the R session with the following clean-up command. 
Warning the following command will remove everything in the working environment 

```{r cleanslate}
rm(list=ls())
```

Let us treat R as a simple calculator

```{r simpleaddition}
2 * 3  #[1] 6
```

Note if R doesnt recognize a code, it displays a + sign and waits for you 

```{r incomplete}
# Uncomment the following line and run using the green arrow to the right
# 2 *
```

```{r error}
# uncomment the following line and run it
 # 10 % 5 
```

### Objects in R

Note that in R everything is treated as an Object. 

Let us start with some basic commands.

```{r assignment}
# non-standard dont use the assignment operator <- 
a = 1  

a <- 1
```

Note that now object a contains number 1. 
Also note that RStudio environment pane now displays the variable

Let us now talk about what are acceptable and not acceptable variable names. 

<pre> 
  No      Yes
  ==      ===
 1run     run1 
  b$       b 
  mean   ^mean (doable but let us avoid it)
  !var1    var1
</pre>

```{r assignment1}
var1 <- 2 
var1 <- 100 #overwrite 
```

```{r morebasiccmds}
ls() # to see what variables you have so far
list.files()  # to see the list of files

currdir <- getwd() # to see where you are
setwd(currdir) # not going anywhere
getwd()   # check 
```

### Basic operations in R

Let us do some data analysis with a die? 

```{r die}
die <- 1:6  # 6 sided die

# let us do some math with die 
# notice the vector math

die + 2

die/2

die * die


# recycling

die + 1:2 


# function 
sum(die)

# Help

# ?sum  # quick way to explore the function arguments

mean(die)
```

### Constants in R 

```{r constants}
# stored constants
pi

# do sume calculations with the constants
# let us round it 

round(pi, 2)

# let us find the arguments of funtion round
args(round)
args(plot)

```

### Drawing samples is an important activity in Statistics

```{r function}
args(sample)

# sampling function in R

# let us sample a die by rolling 3 times

# uncomment (remove the #) and run the following line 
# sample(die, size = 7)

# What was the problem? 

# Read the help page or args function on sample 
# to figure out

#  ?sample

```

```{r roll-a-die}
p <- c(rep(0.01,5), 0.95)
p
sample(x = die, size = 10, replace = TRUE, prob = p)

# no argument name 
sample(die, 10, replace = TRUE)
sample(die, 10, TRUE) #  define them explicitely
sample(x = die, size = 6, replace = TRUE)
```

```{r Multiple-Rolls}
# multiple rolls 
# First roll
die
sample(x = die, size = 2, replace = TRUE) # def prob
# one more roll
sample(x = die, size = 2, replace = TRUE) # def prob
```

### Random numbers

Let us see how to generate random numbers. 

```{r uniformrandom}
unif_rand <- runif(10); unif_rand
args(round)

#### Note the digits = 0; it is set to zero and optional
round(unif_rand, digits = 2)
```

### Functions (like subroutines in Fortran)

Function constructor (like C++) has three parts

 * name 
 * body (of code)
 () arguments 

The function structure looks like the following: function() {  ### }

```{r myfunction1}
dist <- function(a, b) {
  distsq <- sum( (b - a)^2 )
  sqrt(distsq)
}

a <- c(1,2)
b <- c(3,3)

dist(a,b)

```

We can use RStudio to turn the following line into a function and call it  dist2
We can create is just using a few click in RStudio.

Let us do the following steps: 
1) Select 253-254 lines by highlighting 
2) After selecting the lines, do the following, 
  Code --> "Extract Function" and give it a name, my_dist
3) Execute the code 
4) test it with 
   a <- c(1,1,2)
   b <- c(2,2,3)
   my_dist(a,b)

```{r extract-function}
    distsq <- sum( (b - a)^2 )
    sqrt(distsq)
```

### R-packages

```{r rforest}
#install.packages("randomForest")
#library(randomForest)
# update.packages(c("randomForest","tidyverse"))
# update R after update.packages 
```

Let us look at a linear regression example

```{r lr}
## Annette Dobson (1990) "An Introduction to Generalized Linear Models".
## Page 9: Plant Weight Data.
ctl <- c(4.17,5.58,5.18,6.11,4.50,4.61,5.17,4.53,5.33,5.14)
trt <- c(4.81,4.17,4.41,3.59,5.87,3.83,6.03,4.89,4.32,4.69)
group <- gl(2, 10, 20, labels = c("Ctl","Trt"))
weight <- c(ctl, trt)
lm.D9 <- lm(weight ~ group)
lm.D90 <- lm(weight ~ group - 1) # omitting intercept
lm.D9
plot(group,weight, col = "salmon")
```

### Export an object

Save an object to a file 

```{r saveobject}
mtcars
getwd()
write.csv(mtcars, file = "mtcars.csv")
```

### Debugging basics 

Debugging example. Let us define a function called mysum. 

```{r msum}
mysum <- function(a,b) {
  sum( (bb - a)^2 )
}

dist <- function(a, b) {
  distsq <- mysum(a,b)
  sqrt(distsq)
}

# use of RStudio bebug or traceback function
# dist(a,b)
```

use of browser function in debug use the 
control buttons to step through the function 

```{r traceback}
dist2 <- function(a, b) {
  browser()
  distsq <- mysum(a,b)
  sqrt(distsq)
}

## Uncomment and run the following two commands to understand
## how to debug codes

# dist2(a,b)
# 
# debug(dist2(a,b))
```

### Data types
 * six basic types of atomic vectors:
 * doubles, integers, characters, logicals, complex, and raw
 * last two are not very important (see one example below) and we will not discuss further

```{r complex}
complex(1,2)
raw(length = 2)
```


Let us explore the data types  with examples

```{r data-types}
ia <- 1L 
ia1 <- 1
class(ia)
typeof(ia) # internal type or storage mode of any object

class(ia1)
typeof(ia1)

ra <- runif(10)
typeof(ra)
class(ra)

ca <- c("Tom", "cat")
typeof(ca)
class(ca)
```

What is an attribute?

```{r attribute}
x <- cbind(a = 1:3, pi = pi) # simple matrix with dimnames
x
attributes(x)

```

### More on standard R functions

```{r }
matrix(1:6, 3, 2)
args(matrix)

# Density function 
# density function at 5 for a unit normal distribution
dnorm(1,0,5)

dnorm(2, mean = 2, sd = 3)
dnorm(0, mean = 0, sd = 1)

pnorm(0, mean = 0, sd = 1)

```
```{r histogram1}
# Data from UsingR package
# to download do the following
# install.packages("UsingR")
# data(father.son,package="UsingR")

father.son <- read.csv("https://raw.githubusercontent.com/ravichas/bioinformatics/main/data/father.son_UsingR.csv")
rn <- rnorm(1000,0,1)
hist( rn, col = "green" )

x <- father.son$fheight
mean(x)
sd(x)
 hist(x,xlab="Height (in inches)",main="Adult men heights", col = "pink")
  abline(v = 75, col = "blue",lwd = 2) 

```

```{r basics1}
# if trees object is a matrix then then 
# figure from pairs can be imagined to a graphical view of a 
# matrix

#take advantage of system functions
pairs(trees, col = "brown")
```

```{r basics2}
# let us roll a die and sum up the numbers
# first time
results <- sample(x = 1:6, 3) 
results
sum(results)

# second time 
results <- sample(x = 1:6, 3) 
sum(results)

# We do not want to repeat the same commands all the time
# let us put them in a function

roll <- function() {
  die <- 1:6 
  outcome <- sample(x = die, size = 2, replace = TRUE )
#  print(outcome)
  sum(outcome) # returns when you run the function
}

roll()
roll()

```
```{r basics3}
library(tidyverse)
manyrolls <- replicate(10000, roll())
hist(manyrolls, col = "green")

qplot(manyrolls, geom = "histogram")

roll <- function() { 
   die <- 1:6
   dice <- sample(die, size = 2, replace = TRUE)
   sum(dice)
}

roll1 <- function( s = 2) {
   die <= 1:6
   dice <- sample(x = die, size = s, replace = TRUE) # def prob
  sum(dice)
}
```

### How to let RStudio create 

```{r createfunction}
# Turn the following lines into a function in RStudio
# call this a function, myroll. Here are the steps:
#  1) select lines, 450-451 and uncomment them using code --> uncomment Lines
#  2) use ctrl-Alt-X to create a function (works for windows)
#  3) give a function name, myroll  

# after step 3, you should see something like the following 
# 
 
 # myroll <- function(die, s) {
 #   dice <- sample(x = die, size = s, replace = TRUE) # def prob
 #   sum(dice)
 # }
``` 

```{r more-die-options}
# dice <- sample(x = die, size = s, replace = TRUE)
# sum(dice)
 
die <- 1:6
s <- 3
# myroll(die, s )

random <- sample(1:52, size = 52)
random

```

```{r get_symbols}
get_symbols <- function() {
  wheel <- c("DD", "7", "BBB", "BB", "B", "C", "0")
  sample(wheel, size = 3, replace = TRUE,
         prob = c(0.03, 0.03, 0.06, 0.1, 0.25, 0.01, 0.52))
}
get_symbols()

```
